### 2019-11-18 Manuscript_short.Rmd
latest manuscript edits for PLOS Biology

---
bibliography: manuscript.bib
csl: systematic-biology.csl
site: "bookdown::bookdown_site"
output:
  bookdown::pdf_document2:
    toc: FALSE
indent: TRUE
monofont: "Times New Roman"
fontsize: 12pt
header-includes:
  - \usepackage[left]{lineno}
  - \linenumbers
  - \usepackage{setspace}
  - \doublespacing
---

@@

Predators have evolved dedicated body parts to capture and subdue prey. As different predators specialize on distinct prey taxa, their tools for prey capture diverge into a variety of adaptive forms. Studying the evolution of predation is facilitated by a predator clade with structures used exclusively for prey capture and with significant morphological variation. Siphonophores, a clade of colonial cnidarians, satisfy these criteria particularly well, capturing prey with their tentilla (tentacle side branches). Earlier work has shown that extant siphonophore diets correlate with the different morphologies and sizes of their tentilla and nematocysts. We hypothesize that evolutionary specialization on different prey types has driven the phenotypic evolution of these characters. To test this hypothesis, we: (1) measured multiple morphological traits from fixed siphonophore specimens using microscopy and high-speed video techniques, (2) mapped these data to a phylogenetic tree of 45 species, and (3) analyzed the evolutionary associations between siphonophore nematocyst characters and prey type data from the literature. Our results show that siphonophore tentillum structure has strong evolutionary associations with prey type and size specialization, and suggest that shifts between prey-type specializations are linked to shifts in tentillum and nematocyst size and shape. We found that predatory specialists can evolve into generalists, and that specialists on one prey type have directly evolved into specialists on other prey types. Thus, the evolutionary history of tentilla shows that siphonophores are an example of ecological niche diversification via morphological innovation and evolution. This study contributes to understanding how morphological evolution has shaped present-day oceanic food webs.

@@

##   {-}

  Most animal predators have characteristic biological tools that they use to capture and subdue prey. Raptors have claws and beaks, snakes have fangs, wasps have stingers, and cnidarians have nematocyst-laden tentacles. The functional morphology of these structures tend to be finely attuned to their ability to successfully capture specific prey [@schmitz2017predator]. Long-term adaptive evolution in response to the defense mechanisms of the prey (*e.g.* avoidance, escape, protective barriers) leads to modifications that can counter those defenses The more specialized the diet of a predator is, the more specialized its tools need to be to meet the specific challenges posed by the prey. Understanding the relationships between predatory specializations and morphological specializations is necessary to contextualize the phenotypic diversity of predators, and to quantify the importance of ecological diversification in generating this diversity.

  Siphonophores (Cnidaria : Hydrozoa) are a clade of organisms bearing modular structures that are exclusively used for prey capture: the tentilla (Fig. \@ref(figure1)). These present a significant morphological variation across species [@mapstone2014global] (Fig. \@ref(figure2)), which makes them an ideal system to study the relationships between functional traits and prey specialization. A siphonophore is a colony bearing many feeding polyps (Fig. \@ref(figure1)), each with a single tentacle, which branches into several tentilla carrying the functional cnidocytes (specialized neural cells carrying nematocysts, the stinging capsules). Unlike most other cnidarians, siphonophores carry their tentacle nematocysts in extremely complex and organized batteries [@skaer1988formation] built into their tentilla. While nematocyst batteries and clusters in other cnidarians are simple static scaffolds for cnidocytes, siphonophore tentilla have their own reaction mechanism, triggered upon encounter with prey. When it fires, a tentillum undergoes an extremely fast conformational change that wraps it around the prey, maximizing the surface area of contact for nematocysts to fire on the prey [@Mackie:1987uy]. In addition, some species have elaborate fluorescent and bioluminescent lures on their tentilla to attract prey with aggressive mimicry [@purcell1980influence;@haddock2005bioluminescent;@haddock2015fluorescent]. Siphonophores bear four major nematocyst types in their tentacles and tentilla. The largest type, heteronemes, have open-tip tubules characterized by bearing a distinctly wider spiny shaft at the proximal end of the everted tubule. These are typically found flanking the proximal end of the cnidoband. The most abundant type, haplonemes, have no distinct shaft, but similarly to heteronemes, their tubules have open tips and can be found in the cnidoband. Both heteronemes and haplonemes bear short spines along the tubule and can be toxic and penetrate the surface of some prey types. In the terminal filament, siphonophores bear two other types of nematocysts, characterized by their adhesive function, closed tip tubules, and lack of spines on the tubule. These are the desmonemes (a type of adhesive coiled-tubule spironeme), and rhopalonemes (a siphonophore-exclusive nematocyst type with wide tubules).

![\label{figure1} Siphonophore anatomy. A - *Nanomia* sp. siphonophore colony (photo by Catriona Munro). B,C - Illustration of a *Nanomia* colony, gastrozooid, and tentacle (by Freya Goetz). D - *Nanomia* sp. Tentillum illustration and main parts. E - Differential interference contrast micrograph of the tentillum illustrated in D. F - Nematocyst types (illustration reproduced with permission from Mapstone 2014), hypothesized homologies, and locations in the tentillum. Undischarged to the left, discharged to the right.](Figures_final/figure1.jpg)

  Many siphonophore species inhabit the deep pelagic ocean, which spans from ~200m to the oceanic seafloor. This habitat has fairly homogeneous physical conditions and stable zooplankton abundances and composition [@robison2004deep]. With a relatively predictable prey availability, ecological theory would predict evolution to drive coexisting siphonophore lineages towards specialization, increasing their feeding efficiencies and reducing interspecific competition [@simpson1944tempo;@hardin1960competitive;@hutchinson1961paradox]. If this prediction holds true, we expect the prey capture apparatus morphologies of siphonophores to diversify with the evolution of increased specialization on a variety of prey types in different siphonophore lineages.

Specialization is often thought to be an evolutionary ‘dead end’, meaning that specialized lineages are unlikely to evolve into generalists or to shift the resource for which they are specialized [@futuyma1988evolution]. However, recent studies have found that interspecific competition can favor the evolution of resource generalism [@stireman2005evolution;@johnson2009competition] and resource switching [@hoberg2008macroevolutionary]. Here we examine three alternative hypotheses on siphonophore trophic specialization: (1) predatory specialists evolved from generalist ancestors; (2) predatory specialists evolved from ancestral predators which specialized on a different resource, switching their primary prey type; and (3) predatory generalists evolved from specialist ancestors.

@@

  The few pioneering studies that have addressed the relationships between tentilla and diet suggest that siphonophores are a robust system for the study of predatory specialization via morphological diversification. [@purcell1984functions] and [@purcell1988correlation] showed clear relationships between diet, tentillum, and nematocyst characters in co-occurring epipelagic siphonophores. These correlations, while studied for a small subset of extant epipelagic siphonophore species, might be generalizable to all siphonophores. We hypothesize that these relationships reflect correlated evolution between prey selection and tentillum (and nematocyst) traits. Furthermore, we hypothesize that with an extensive characterization of tentilla morphology, we can generate hypotheses about the diets of understudied siphonophore species.

  In this study, we characterize the morphological diversity of tentilla and their nematocysts across a broad variety of shallow and deep sea siphonophore species using modern imaging technologies, we expand the phylogenetic tree of siphonophores by combining a broad taxon sampling of ribosomal gene sequences with a transcriptome-based backbone tree, and we explore the evolutionary histories and correlations among diet, tentillum, and nematocyst characters.

@@

@@

*Phylogeny* – Only 5 nodes (blue dots in Figure \@ref(figure4)) in the unconstrained inference were incongruent with the [@munro2018improved] transcriptome tree, and these were constrained during estimation of the 18S+16S tree. The topology of the constrained tree presented here (Fig. \@ref(figure4)) is congruent with the resolved nodes in [@dunn2005molecular] and [@munro2018improved].

![\label{figure4} Bayesian time-tree built from 18S + 16S concatenated sequences. Branch lengths estimated using relaxed molecular clock. Species names in red indicate replicated representation in the morphology data. Species marked with a red asterisk were recorded using high speed video. Newly accessioned 16S data was used for species marked with a blue circle. Nodes labeled with Bayesian posteriors (BP). Green circles indicate BP = 1. Blue circles indicate nodes constrained to be congruent with (Munro *et al.* 2018). This phylogeny expands the taxon sampling of the (Munro *et al.* 2018) transcriptome tree and is congruent with it. Tips with black squares indicate the species with transcriptomes used in (Munro *et al.* 2018). Tips with grey squares indicate genus-level correspondence to taxa included in (Munro *et al.* 2018). The main clades are labeled: in black for described taxonomic units, and in grey for operational phylogenetic designations.](Figures_final/figure4.png){height=600px}

@@

*Evolutionary dynamics between diet and tentillum morphology* – Reconstructions of feeding guilds shows that generalism is not likely to be ancestral, and it appears to have evolved at least two times independently (Fig. \@ref(figure5)). Large crustacean specialists evolve into generalists twice independently, supporting hypothesis 3. Feeding guild specializations have shifted from an alternative ancestral state at least five times, supporting hypothesis 2. Copepod specialization and fish specialization evolved twice, and ostracod specialization evolved at least once.

@@

  Phylogenetic regressions of continuous characters against prey selectivity data produced additional insights. Fish selectivity is associated with increased number of heteronemes per tentillum, increased roundness of nematocysts (desmonemes and haplonemes), larger heteronemes, reduced heteroneme/cnidoband length ratios, smaller rhopalonemes, lower haploneme SA/V ratios, and increased size of the cnidoband, elastic strand, pedicle and tentacle widths. Decapod-selective diets were associated with increasing cnidoband size and coiledness, haploneme row number, elastic strand width, and heteroneme number. Copepod-selective diets evolved in association with smaller heteroneme and total nematocyst volumes, smaller cnidobands, rounder rhopalonemes, elongated heteronemes, narrower haplonemes with higher SA/V ratios, and smaller heteronemes, tentacles, pedicles and elastic strands. Selectivity for ostracods was associated with reductions in size and number of heteroneme nematocysts, reductions in cnidoband size, number of haploneme rows, heteroneme number, and cnidoband coiledness. Heteroneme length and elongation also correlated negatively with chaetognath selectivity.

@@

*Generating dietary hypotheses using tentillum morphology* – The discriminant analysis of principal components for feeding guild (7 principal components, 4 discriminants) produced 100% discrimination, and the highest loading contributions were found for the characters (ordered from highest to lowest): Involucrum length, heteroneme volume, heteroneme number, total heteroneme volume, tentacle width, heteroneme length, total nematocyst volume, and heteroneme width (SM21). We used the predictions from this discriminant function to generate hypotheses about the feeding guild of 45 species in our morphological data (SM2). This projection predicts that two other *Apolemia* species may also be gelatinous prey specialists like *Apolemia rubriversa*, and that *Erenna laciniata* may be a fish specialist like *Erenna richardi*.

Table 2. Discriminant analysis of principal components for the presence of specific prey types using the morphological data. Top quartile variable (character) contributions to the linear discriminants are ordered from highest to lowest. Logistic regressions and GLMs were fitted to predict prey type presence and selectivity respectively. The sign of the slope of each predictor is reported, marked with an asterisk if significant (p value < 0.05), and highlighted grey if it differs between prey presence in diet and prey selectivity. Pseudo-R^2^ (%) approximates the percent variance explained by the model.

@@

*Phenotypic integration of the tentillum* – The quantitative characters we measured from tentilla and their nematocysts are highly correlated. The results indicate that the dimensionality of tentillum morphology is low, that many traits are associated with size, but that nematocyst arrangement and shape are independent of it. The variance covariance matrices (SM36-38) are congruent with the abundant positive correlations observed among simple measurement characters in SM3. However, this analysis reveals more clearly the diagonal blocks that constitute the evolutionary modules, such as the heteroneme block, the terminal filament nematocyst block, and the cnidoband-pedicle-tentacle block. These results were not very sensitive to transformation of inapplicable states and taxon sampling. When we compared the rate covariance terms between characters across the different feeding guild regimes (SM41), we found that half (48%) of the character pairs presented distinct correlation coefficients across different regimes, indicating that the mode of phenotypic integration may also shift with trophic niche. When contrasting the regime-specific rate correlation matrices to the whole-tree matrix, we were able to identify the character dependencies that are unique to each predatory niche (SM42).

Under the majority of SIMMAP outcomes, large crustaceans specialists are the first regime to appear, and other regimes evolve in a shift from this ancestral specialization. Compared to the rate correlation matrix estimated over the whole tree, large crustacean specialists present strong negative correlations between haploneme elongation and heteroneme size, and between rhopaloneme elongation and tentillum size, as well as with involucrum length. With the appearance of generalists (*Forskalia* and the *Agalma-Athorybia* clade), terminal filament nematocyst (desmonemes and rhopalonemes) sizes became negatively correlated with the sizes of most characters, meaning that as some tentilla became larger, their individual terminal nematocysts became smaller, observed to the extreme in *Agalma*. In addition, heteroneme and rhopaloneme elongation became positively correlated with cnidoband size. When large crustacean specialists switched to small crustacean prey in *Cordagalma* and calycophorans, haploneme size became inversely correlated with heteroneme elongation, which in turn developed a strong positive relationship with tentillum size. In other words, as tentilla get smaller in this group, heteronemes get shorter and haplonemes get larger. The extremes of this gradient can be seen in *Cordagalma* and *Hippopodius*. With the evolution of fish prey specialization in cystonects and within Clade B, haploneme elongation became negatively correlated with heteroneme elongation (signal driven by Clade B, since cystonects lack tentacular heteronemes), and the surface area to volume ratio of haploneme nematocysts switched from a strong negative relationship with cnidoband size (found in every other regime) to a positively correlation. Gelatinous specialization, albeit appearing only once in our tree, also carries a unique signature in character rate correlation shifts, with an increase in the strength of the correlation between heteroneme shape and shaft width, consistent with the appearance of birrhopaloid nematocysts with swollen shafts that are likely effective at anchoring gelatinous tissue (see reference to Narcomedusae nematocysts in [@purcell1988correlation]).

@@

The phylogenetic positions of the main categorical character shifts were reconstructed using stochastic character mapping (SM), and summarized in \@ref(figure7). Haploneme nematocysts are likely ancestrally present in siphonophore tentacles, since they are present in the tentacles of many other hydrozoans. Haplonemes first diverged into spherical isorhizas of 2 size classes in Cystonectae, and elongated anisorhizas of one size class in Codonophora. Haplonemes were likely lost in the tentacles of *Apolemia*, but spherical isorhizas are retained in other *Apolemia* tissues [@siebert2013re]. Similarly, while heteronemes exist in other tissues of cystonects, they appear in the tentacles of codonophorans exclusively, as birhopaloids in *Apolemia*, ancestral stenoteles in eucladophoran physonects, and microbasic mastigophores in calycophorans.

  Eucladophora (the clade containing Pyrostephidae, Euphysonectae, and Calycophorae, see main text Fig. 3) encompasses most of the extant Siphonophore species (178 of 186) other than Cystonects and *Apolemia*. Innovations evolved in the stem of this group include spatially segregated heteroneme and haploneme nematocysts, terminal filaments, and elastic strands (Fig. \@ref(figure7)). Pyrostephids evolved a unique bifurcation of the axial gastrovascular canal of the tentillum known as the "saccus" [@totton1965synopsis]. The stem to the clade Tendiculophora (clade containing Euphysonectae and Calycophorae, see main text Fig. 3) subsequently acquired further novelties such as the desmoneme and rhopaloneme (acrophore subtype ancestral) nematocysts on the terminal filament (Fig. \@ref(figure7)), which bears no other nematocyst type (main text Fig. 1). These are arranged in sets of 2 parallel rhopalonemes for each single desmoneme [@skaer1988formation;@skaer1991remodelling]. The involucrum is an expansion of the epidermal layer that can cover part or all of the cnidoband (main text Fig. 2). This structure, together with differentiated larval tentilla, appeared in the stem branch to Clade A physonects. Calycophorans evolved novelties such as larger desmonemes at the distal end of the cnidoband, pleated pedicles with a “hood” (here considered homologous to the involucrum) at the proximal end of the tentillum, anacrophore rhopalonemes, and microbasic mastigophore-type heteronemes. While calycophorans have diversified into most of the extant described siphonophore species (108 of 186), their tentilla have not undergone any major categorical gains or losses since their most recent common ancestor. Nonetheless, they have spreaded over a broad span of variation in nematocyst and cnidoband sizes.

*Evolution of nematocyst shape* – Haploneme nematocyst evolution has been mainly driven by a single large shift towards elongation in Tendiculophora, which contains the majority of described siphonophore species other than Cystonects, *Apolemia*, and Pyrostephidae. There is one secondary return to more oval, less elongated haplonemes in *Erenna*, but it does not reach the sphericity present in Cystonectae or Pyrostephidae (Fig. \@ref(figure10). Heteroneme evolution presents a less discrete evolutionary history, where Tendiculophora evolved more elongate heteronemes, but the difference between theirs and other siphonophores is much smaller than the variation in shape within Tendiculophora, bearing no phylogenetic signal. In this clade, evolution of heteroneme shape has diverged in both directions, and there is no correlation with haploneme shape (Fig. \@ref(figure10)), which has remained fairly constant (elongation between 1.5 and 2.5).

![\label{figure10} Phylomorphospace showing haploneme and heteroneme elongation (log scaled). Orange area delimits rod-shaped haplonemes, blue area covers oval and round shaped haplonemes. Smaller dots and lines represent phylogenetic relationships and ancestral states of internal nodes under BM. Species nodes in red were manually added to the plot. Cystonects have no tentacle heteronemes and are projected onto the haploneme axis. Apolemiids have no tentacle haplonemes and are projected onto the heteroneme axis. Colored branches and nodes correspond to BAMM regimes of accelerated haploneme shape (green) and heteroneme shape (violet) evolution.](Figures_final/figure10.jpg)

>>>>

---
bibliography: manuscript.bib
csl: plos-biology.csl
site: "bookdown::bookdown_site"
output:
  bookdown::pdf_document2:
    toc: FALSE
indent: TRUE
monofont: "Times New Roman"
fontsize: 12pt
header-includes:
  - \usepackage[left]{lineno}
  - \linenumbers
  - \usepackage{setspace}
  - \doublespacing
---

@@

Predators have evolved dedicated body parts to capture and subdue prey. As a clade of predators specialize in distinct prey taxa, their tools for prey capture diversify into a variety of adaptive forms. Studying the evolution of predation is facilitated by a predator clade with structures used exclusively for prey capture and with significant morphological variation. Siphonophores, a clade of colonial cnidarians, satisfy these criteria particularly well, capturing prey with their tentilla (tentacle side branches). Earlier work has shown that extant siphonophore diets correlate with the different morphologies and sizes of their tentilla and nematocysts. We hypothesize that evolutionary specialization on different prey types has driven the phenotypic evolution of these characters. To test this hypothesis, we: (1) measured multiple morphological characters from siphonophore tentacle specimens, (2) mapped these data to a phylogenetic tree of 45 species, and (3) analyzed the evolutionary associations between morphological characters and prey type data from the literature. Our results show that the siphonophore tentillum morphology has strong evolutionary associations with prey type specialization, and suggest that shifts between prey-type specializations are linked to shifts in tentillum and nematocyst size and shape. We found that predatory specialists can evolve into generalists and that specialists on one prey type have directly evolved into specialists on other prey types. With changes in trophic niche regimes, both trait optima and trait correlation patterns showed significant shifts. Thus, the evolutionary history of tentilla shows that siphonophores are an example of ecological niche diversification via morphological innovation and evolution. This study contributes to understanding how morphological evolution has shaped present-day oceanic food webs.

@@

## Introduction {-}

  Most animal predators have characteristic biological tools that they use to capture and subdue prey. Raptors have claws and beaks, snakes have fangs, wasps have stingers, and cnidarians have nematocyst-laden tentacles. The functional morphology of these structures tends to be finely attuned to their ability to successfully capture specific prey [@schmitz2017predator]. Long-term adaptive evolution in response to the defense mechanisms of the prey (*e.g.*, avoidance, escape, protective barriers) leads to modifications that can counter those defenses. The more specialized the diet of a predator is, the more specialized its tools need to be to meet the specific challenges posed by the prey. Understanding the relationships between predatory and morphological specializations is necessary to contextualize the phenotypic diversity of predators, and to quantify the importance of ecological diversification in generating this diversity.

  Siphonophores (Cnidaria: Hydrozoa) are a clade of organisms bearing modular structures that are exclusively used for prey capture: the tentilla (Fig. \@ref(figure1)). These present significant morphological variation across species [@mapstone2014global] (Fig. \@ref(figure2)), which makes them an ideal system to study the relationships between functional traits and prey specialization. A siphonophore is a colony bearing many feeding polyps (Fig. \@ref(figure1)), each with a single tentacle, which branches into several tentilla carrying the functional cnidocytes (specialized neural cells carrying nematocysts, the stinging capsules). Unlike most other cnidarians, siphonophores carry their tentacle nematocysts in extremely complex and organized batteries [@skaer1988formation] built into their tentilla (tentacle side branches). While nematocyst batteries and clusters in other cnidarians are simple static scaffolds for cnidocytes, siphonophore tentilla have their own reaction mechanism, triggered upon encounter with prey. When it fires, a tentillum undergoes an extremely fast conformational change that wraps it around the prey, maximizing the surface area of contact for nematocysts to fire on the prey [@Mackie:1987uy]. In addition, some species have elaborate fluorescent and bioluminescent lures on their tentilla to attract prey with aggressive mimicry [@purcell1980influence;@haddock2005bioluminescent;@haddock2015fluorescent]. Siphonophores bear four major nematocyst types in their tentacles and tentilla (Fig. \@ref(figure1)F).. The largest type, heteronemes, have open-tip tubules characterized by bearing a distinctly wider spiny shaft at the proximal end of the everted tubule. These are typically found flanking the proximal end of the cnidoband. The most abundant type, haplonemes, have no distinct shaft, but similarly to heteronemes, their tubules have open tips and can be found in the cnidoband. Both heteronemes and haplonemes bear short spines along the tubule and can be toxic and penetrate the surface of some prey types. In the terminal filament, siphonophores bear two other types of nematocysts, characterized by their adhesive function, closed tip tubules, and lack of spines on the tubule. These are the desmonemes (a type of adhesive coiled-tubule spironeme), and rhopalonemes (a siphonophore-exclusive nematocyst type with wide tubules).

![\label{figure1} Siphonophore anatomy. A - *Nanomia* sp. siphonophore colony (photo by Catriona Munro). B, C - Illustration of a *Nanomia* colony, gastrozooid, and tentacle (by Freya Goetz). D - *Nanomia* sp. Tentillum illustration and main parts. E - Differential interference contrast micrograph of the tentillum illustrated in D. F - Nematocyst types (illustration reproduced with permission from Mapstone 2014), hypothesized homologies, and locations in the tentillum. Undischarged to the left, discharged to the right.](Figures_final/figure1.jpg)

  Many siphonophore species inhabit the deep pelagic ocean, which spans from ~200m to the oceanic seafloor. This habitat has fairly homogeneous physical conditions and stable zooplankton abundances and composition [@robison2004deep]. With a relatively predictable prey availability, ecological theory would predict evolution to drive coexisting siphonophore lineages towards specialization, increasing their feeding efficiencies and reducing interspecific competition [@simpson1944tempo;@hardin1960competitive;@hutchinson1961paradox]. If this prediction holds true, we expect the prey capture apparatus morphologies of siphonophores to diversify with the evolution of increasing specialization on a variety of prey types in different siphonophore lineages.

  Specialization has been thought to be an evolutionary ‘dead-end’, meaning that specialized lineages are unlikely to evolve into generalists or to shift the resource for which they are specialized [@futuyma1988evolution]. However, recent studies have found that interspecific competition can favor the evolution of resource generalism [@stireman2005evolution; @johnson2009competition; @nosil2002transition] and resource switching [@hoberg2008macroevolutionary; @hardy2014specialization]. Here we examine three alternative hypotheses on siphonophore trophic specialization: (1) predatory specialists evolved from generalist ancestors; (2) predatory specialists evolved from ancestral predators which specialized on a different resource, switching their primary prey type; and (3) predatory generalists evolved from specialist ancestors.

@@

@@

  The few pioneering studies that have addressed the relationships between tentilla and diet suggest that siphonophores are a robust system for the study of predatory specialization via morphological diversification. [@purcell1984functions] and [@purcell1988correlation] showed clear relationships between diet, tentillum, and nematocyst characters in co-occurring epipelagic siphonophores. These correlations, while studied for a small subset of extant epipelagic siphonophore species, might be generalizable to all siphonophores. We hypothesize that these relationships reflect correlated evolution between prey selection and tentillum (and nematocyst) traits. Furthermore, we hypothesize that with extensive characterization of tentilla morphology, we can generate hypotheses about the diets of understudied siphonophore species.

  In this study, we characterize the morphological diversity of tentilla and their nematocysts across a broad variety of shallow and deep-sea siphonophore species using modern imaging technologies, we expand the phylogenetic tree of siphonophores by combining a broad taxon sampling of ribosomal gene sequences with a transcriptome-based backbone tree, and we explore the evolutionary histories and correlations among diet, tentillum, and nematocyst characters.

@@

*Phylogeny* – We built upon the published siphonophore transcriptome phylogeny [@munro2018improved] to produce a tree with extensive taxonomic representation using ribosomal genes. Only 5 nodes (blue dots in Figure \@ref(figure4)) in the unconstrained inference were incongruent with the [@munro2018improved] transcriptome tree, and these were constrained during estimation of the 18S+16S tree. The topology of the constrained tree presented here (Fig. \@ref(figure4)) is congruent with the resolved nodes in [@dunn2005molecular] and [@munro2018improved]. However, with the unprecedented inclusion of *Stephanomia amphytridis*, this phylogeny reveals a novel phylogenetic relationship between the genus *Erenna* and *Stephanomia*.

![\label{figure4} Bayesian time-tree built from 18S + 16S concatenated sequences. Branch lengths estimated using a relaxed molecular clock. Species names in red indicate replicated representation in the morphology data. Species marked with a red asterisk were recorded using high-speed video. Newly accessioned 16S data was used for species marked with a blue circle. Nodes labeled with Bayesian posteriors (BP). Green circles indicate BP = 1. Blue circles indicate nodes constrained to be congruent with (Munro *et al.* 2018). This phylogeny expands the taxon sampling of the (Munro *et al.* 2018) transcriptome tree and is congruent with it. Tips with black squares indicate the species with transcriptomes used in (Munro *et al.* 2018). Tips with grey squares indicate genus-level correspondence to taxa included in (Munro *et al.* 2018). The main clades are labeled: in black for described taxonomic units, and in grey for operational phylogenetic designations.](Figures_final/figure4.png){height=600px}

@@

*Evolutionary associations between diet and tentillum morphology* – Reconstructions of feeding guilds show that generalism is not likely to be ancestral, and it appears to have evolved at least two times independently (Fig. \@ref(figure5)). Large crustacean specialists evolve into generalists twice independently, supporting hypothesis 3. Feeding guild specializations have shifted from an alternative ancestral state at least five times, supporting hypothesis 2. Copepod specialization and fish specialization evolved twice, and ostracod specialization evolved at least once.

@@

  Phylogenetic regressions of continuous characters against prey selectivity data produced additional insights. Fish selectivity is associated with increased number of heteronemes per tentillum, increased roundness of nematocysts (desmonemes and haplonemes), larger heteronemes, reduced heteroneme/cnidoband length ratios, smaller rhopalonemes, lower haploneme SA/V ratios, and increased size of the cnidoband, elastic strand, pedicle and tentacle widths. Decapod-selective diets were associated with increasing cnidoband size and coiledness, haploneme row number, elastic strand width, and heteroneme number. Copepod-selective diets evolved in association with smaller heteroneme and total nematocyst volumes, smaller cnidobands, rounder rhopalonemes, elongated heteronemes, narrower haplonemes with higher SA/V ratios, and smaller heteronemes, tentacles, pedicles, and elastic strands. Selectivity for ostracods was associated with reductions in size and number of heteroneme nematocysts, reductions in cnidoband size, number of haploneme rows, heteroneme numbers, and cnidoband coiledness. Heteroneme length and elongation also correlated negatively with chaetognath selectivity (SM21).

@@

*Generating dietary hypotheses using tentillum morphology* – The discriminant analysis of principal components for feeding guild (7 principal components, 4 discriminants) produced 100% discrimination between guilds, and the highest loading contributions were found for the characters (ordered from highest to lowest): Involucrum length, heteroneme volume, heteroneme number, total heteroneme volume, tentacle width, heteroneme length, total nematocyst volume, and heteroneme width (SM21). We used the predictions from this discriminant function to generate hypotheses about the feeding guild of 45 species in our morphological data (SM2). This projection predicts that two other *Apolemia* species may also be gelatinous prey specialists like *Apolemia rubriversa* and that *Erenna laciniata* may be a fish specialist like *Erenna richardi*.

Table 2. Discriminant analysis of principal components for the presence of specific prey types using the morphological data. Top quartile variable (character) contributions to the linear discriminants are ordered from highest to lowest. Logistic regressions and GLMs were fitted to predict prey type presence and selectivity respectively. The sign of the slope of each predictor is reported, marked with an asterisk if significant (p-value < 0.05), and highlighted grey if it differs between prey presence in diet and prey selectivity. Pseudo-R^2^ (%) approximates the percent variance explained by the model.

@@

*Phenotypic integration of the tentillum* – The quantitative characters we measured from tentilla and their nematocysts are highly correlated. The results indicate that the dimensionality of tentillum morphology is low, that many traits are associated with size, but that nematocyst arrangement and shape are independent of it (SM4). The variance-covariance matrices (SM36-38) are congruent with the abundant positive correlations observed among simple measurement characters in SM3. However, this analysis reveals more clearly the diagonal blocks that constitute the evolutionary modules, such as the heteroneme block, the terminal filament nematocyst block, and the cnidoband-pedicle-tentacle block. These results were not sensitive to the transformation of inapplicable states and taxon sampling. When we compared the rate covariance terms between characters across the different feeding guild regimes (SM41), we found that half (48%) of the character pairs presented significantly distinct correlation coefficients across different regimes (SM39), indicating that the mode of phenotypic integration also shifts with trophic niche. When contrasting the regime-specific rate correlation matrices to the whole-tree matrix, we were able to identify the character dependencies that are unique to each predatory niche (SM42).

Under the majority of SIMMAP outcomes, large crustaceans specialists are the ancestral regime, and allß other regimes evolve from this ancestral specialization. Compared to the rate correlation matrix estimated over the whole tree, large crustacean specialists present strong negative correlations between haploneme elongation and heteroneme size, and between rhopaloneme elongation and tentillum size, as well as with involucrum length. With the appearance of generalists (*Forskalia* and the *Agalma-Athorybia* clade), terminal filament nematocyst (desmonemes and rhopalonemes) sizes became negatively correlated with the sizes of most characters, meaning that as some tentilla became larger, their individual terminal nematocysts became smaller, observed to the extreme in *Agalma*. In addition, heteroneme and rhopaloneme elongation became positively correlated with cnidoband size. When large crustacean specialists switched to small crustacean prey in *Cordagalma* and calycophorans, haploneme size became inversely correlated with heteroneme elongation, which in turn developed a strong positive relationship with tentillum size. In other words, as tentilla get smaller in this group, heteronemes get shorter and haplonemes get larger. The extremes of this gradient can be seen in *Cordagalma* and *Hippopodius*. With the evolution of fish prey specialization in cystonects and within Clade B, haploneme elongation became negatively correlated with heteroneme elongation (signal driven by Clade B, since cystonects lack tentacular heteronemes), and the surface area to volume ratio of haploneme nematocysts switched from a strong negative relationship with cnidoband size (found in every other regime) to a positive correlation. Gelatinous specialization, albeit appearing only once in our tree, also carries a unique signature in character rate correlation shifts, with an increase in the strength of the correlation between heteroneme shape and shaft width, consistent with the appearance of birrhopaloid nematocysts with swollen shafts that are likely effective at anchoring gelatinous tissue (see reference to Narcomedusae nematocysts in [@purcell1988correlation]).

@@

The phylogenetic positions of the main categorical character shifts were reconstructed using stochastic character mapping (SM22-30) and summarized in \@ref(figure7). Haploneme nematocysts are likely ancestrally present in siphonophore tentacles since they are present in the tentacles of many other hydrozoans. Haplonemes first diverged into spherical isorhizas of 2 size classes in Cystonectae, and elongated anisorhizas of one size class in Codonophora. Haplonemes were likely lost in the tentacles of *Apolemia* but retained as spherical isorhizas in other *Apolemia* tissues [@siebert2013re]. Similarly, while heteronemes exist in other tissues of cystonects, they appear in the tentacles of codonophorans exclusively, as birhopaloids in *Apolemia*, stenoteles in eucladophoran physonects, and microbasic mastigophores in calycophorans.

  Eucladophora (the clade containing Pyrostephidae, Euphysonectae, and Calycophorae, see Fig. \@ref(figure3)) encompasses most of the extant Siphonophore species (178 of 186) other than Cystonects and *Apolemia*. Innovations evolved in the stem of this group include spatially segregated heteroneme and haploneme nematocysts, terminal filaments, and elastic strands (Fig. \@ref(figure7)). Pyrostephids evolved a unique bifurcation of the axial gastrovascular canal of the tentillum known as the "saccus" [@totton1965synopsis]. The stem to the clade Tendiculophora (clade containing Euphysonectae and Calycophorae, Fig. \@ref(figure3)) subsequently acquired further novelties such as the desmoneme and rhopaloneme (acrophore subtype ancestral) nematocysts on the terminal filament (Fig. \@ref(figure7)), which bears no other nematocyst type (Fig. \@ref(figure1)). These are arranged in sets of 2 parallel rhopalonemes for each single desmoneme [@skaer1988formation;@skaer1991remodelling]. The involucrum is an expansion of the epidermal layer that can cover part or all of the cnidoband (Fig. \@ref(figure2)). This structure, together with differentiated larval tentilla, appeared in the stem branch to Clade A physonects. Calycophorans evolved novelties such as larger desmonemes at the distal end of the cnidoband, pleated pedicles with a “hood” (here considered homologous to the involucrum) at the proximal end of the tentillum, anacrophore rhopalonemes, and microbasic mastigophore-type heteronemes. While calycophorans have diversified into most of the extant described siphonophore species (108 of 186), their tentilla have not undergone any major categorical gains or losses since their most recent common ancestor. Nonetheless, they have spread over a broad span of variation in nematocyst and cnidoband sizes.

*Evolution of nematocyst shape* – Haploneme nematocyst evolution has been mainly driven by a single large shift towards elongation in Tendiculophora, which contains the majority of described siphonophore species other than Cystonects, *Apolemia*, and Pyrostephidae. There is one secondary return to more oval, less elongated haplonemes in *Erenna*, but it does not reach the sphericity present in Cystonectae or Pyrostephidae (Fig. \@ref(figure10)). Heteroneme evolution presents a less discrete evolutionary history, where Tendiculophora evolved more elongate heteronemes, but the difference between theirs and other siphonophores is much smaller than the variation in shape within Tendiculophora, bearing no phylogenetic signal. In this clade, the evolution of heteroneme shape has diverged in both directions, and there is no correlation with haploneme shape (Fig. \@ref(figure10)), which has remained fairly constant (elongation between 1.5 and 2.5).

![\label{figure10} Phylomorphospace showing haploneme and heteroneme elongation (log scaled). Orange area delimits rod-shaped haplonemes, blue area covers oval and round-shaped haplonemes. Smaller dots and lines represent phylogenetic relationships and ancestral states of internal nodes under BM. Species nodes in red were manually added to the plot. Cystonects have no tentacle heteronemes and are projected onto the haploneme axis. Apolemiids have no tentacle haplonemes and are projected onto the heteroneme axis. Colored branches and nodes correspond to BAMM regimes of accelerated haploneme shape (green) and heteroneme shape (violet) evolution.](Figures_final/figure10.jpg)
